---
title: "Systematic Review of Ionized Magnesium"
author: "Sasha Haywood, Chen Zhou"
date: "3/17/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = F}
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
iMg = read_csv("iMg_EXTRACTION_Resultsfinal_Ansu.csv")
iMg = iMg[,c(1, 4, 7, 9, 10, 11, 12)] # 261 rows
iMg = iMg[!is.na(iMg$grp1_base_mean),] # remove 18 rows without means
iMg = iMg[!is.na(iMg$grp1_base_n),] # remove 10 rows without sample sizes
iMg = unite(iMg, group, 2:3, remove = T)

# replace missing sd with sd calculated from se and n
missing.sd = is.na(iMg$grp1_base_sd)
iMg$grp1_base_sd[missing.sd] = with(iMg, grp1_base_se[missing.sd] * 
                                      sqrt(grp1_base_n[missing.sd]))
iMg = iMg[!is.na(iMg$grp1_base_sd),] # remove 22 rows without sd
  
groups = iMg %>%
  group_by(group) %>%
  mutate(n = sum(grp1_base_n)) %>%
  mutate(weighted.mean = grp1_base_mean * grp1_base_n / (n-1)) %>%
  mutate(mean = sum(weighted.mean)) %>%
  mutate(q = (grp1_base_n - 1) * grp1_base_sd^2 + grp1_base_n * grp1_base_mean^2) %>%
  mutate(qc = sum(q)) %>%
  mutate(var = (qc - n * mean^2) / (n-1)) %>%
  mutate(sd = sqrt((qc - n * mean^2) / (n-1)))

report1 = groups[,c(2, 9, 13)]
unique(report1)


######### alternate code 
########## this code leave in the lines that only have mean until overall mean has been calculated
############ and then it removes those lines and recalculates n and calculates overall sd

iMg = read_csv("iMg_EXTRACTION_Resultsfinal_Ansu.csv")
iMg = iMg[,c(1, 4, 7, 9, 10, 11, 12)] # 261 rows
iMg = iMg[!is.na(iMg$grp1_base_mean),] # remove 18 rows without means
iMg = iMg[!is.na(iMg$grp1_base_n),] # remove 10 rows without sample sizes
iMg = unite(iMg, group, 2:3, remove = T)

# replace missing sd with sd calculated from se and n
missing.sd = is.na(iMg$grp1_base_sd)
iMg$grp1_base_sd[missing.sd] = with(iMg, grp1_base_se[missing.sd] * 
                                      sqrt(grp1_base_n[missing.sd]))


groups = iMg %>%
  group_by(group) %>%
  mutate(n = sum(grp1_base_n)) %>%
  mutate(weighted.mean = grp1_base_mean * grp1_base_n / (n-1)) %>%
  mutate(mean = sum(weighted.mean))

groups = groups[!is.na(groups$grp1_base_sd),] # remove 22 rows without sd

groups = groups %>%
  mutate(new_n = sum(grp1_base_n)) %>%
  mutate(q = (grp1_base_n - 1) * grp1_base_sd^2 + grp1_base_n * grp1_base_mean^2) %>%
  mutate(qc = sum(q)) %>%
  mutate(var = (qc - new_n * mean^2) / (n-1)) %>%
  mutate(sd = sqrt((qc - new_n * mean^2) / (n-1)))

report2 = groups[,c(2, 9, 14)]

# compare difference in means, depending on when we remove the lines with missing sd
m1 = unique(report1)
m2 = unique(report2)

# rows with NaN overall sd have only one observation, so overall sd is same as original sd
m1[10,3] = 0.080
m2[10,3] = 0.080
m1[11,3] = 0.055
m2[11,3] = 0.055

m1 = m1[order(m1$group),] 
m2 = m2[order(m2$group),] 

m1 = m1 %>%
  mutate(lower = mean - 1.96 * sd) %>%
  mutate(upper = mean + 1.96 * sd)
m2 = m2 %>%
  mutate(lower = mean - 1.96 * sd) %>%
  mutate(upper = mean + 1.96 * sd)

m1 = separate(m1, group, c("health condition", "metric"), sep = "_")
m2 = separate(m2, group, c("health condition", "metric"), sep = "_")

m1[6,2] = "Serum Mg"
m2[6,2] = "Serum Mg"
```

# Introduction
The end goal of this project is to determine a reference range for ionized magnesium and total magnesium concentrate in people with different health conditions.  The results from a number of studies have been compiled, and our initial objective is to combine the means and standard deviations from those seperate studies.  The initial collection of data includes studies of people with five different health conditions.

 * diabetes 
 * hypertension
 * renal disease
 * cardiovascular disease
 * healthy

For each health condition, there are two measurements.

 * ionized magnesium 
 * total magnesium concentration.

Additionally, the data includes total serum magnesium concentration for healthy individuals.

# The Data
The data set contains one observation for each study/group/condition/measurement combination.  Not all conditions are included in all studies.  For example, there are 24 study/group observations reporting ionized magnesium in people with cardiovascular disease, but only one reporting total magnesium in people with hypertension.  Studies range in size from samples of 6 to 1,652 people.  For each study, we should have sample size, mean, and standard deviation or standard error.  In the cases where the mean or sample size is missing, there is no way to include the observation in a weighted mean.  This includes 28 rows which must be excluded from all calculations.  A number of observations have a reported standard error, but no standard deviation.  In those cases, the standard deviation has been calculated using the formula $s = s_{\bar x} \times \sqrt n$
After making this calculation, there remain 22 observations with no standard deviation.  We have managed this discrepancy in two different ways, explained in the Methods section.

# The Math
The weighted means are calculated using the formula

$$
\bar x_t=\  \sum_{i=1}^{k}\frac{\bar x_i \times n_i}{n_t-1} \; \; \text{where} \; \; n_t = \sum_{i=1}^{k}n_i 
$$

The weighted standard deviations are calculated using the formula

$$
s_t = \sqrt\frac{{\sum_{i=1}^{n}((n_i - 1) \times s_i^2 + n_i \times \bar x_i ^2) - n_i \times \bar x_t^2}}{n_t-1}
$$

# Two Methods
In our first method, we started by removed the observations with missing standard deviations.   We then calculated the weighted mean ($\bar x_t$) and weighted standard deviations ($s_t$) without those observations.

In our second method, we calculated the weighted mean using all observations for which we had both sample size and mean.  We then removed the 22 observations with no standard deviation, recalculated the total sample sizes ($n_t$), and calculated the standard deviation. In our opinion, this is a valid method for analysis and is preferred, as it includes as much information as is available.  Upper and lower bounds are then calculated for all conditions and metrics.

# The Results

\begin{center}

Method 1

\end{center}

```{r, echo = F}
kable(m1, )
```

\begin{center}

Method 2

\end{center}

```{r, echo = F}
kable(m2, )
``` 


